---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r}
# Trying to load dependencies
library(data.table)
library(ggplot2)
```

## Loading and preprocessing the data

```{r}
# Load activities
activities.data.table <- fread("activity.csv")
# Convert dates to POSIX format
activities.data.table[, date := as.POSIXct(strptime(date, format = "%Y-%m-%d"))]
```

## What is mean total number of steps taken per day?

Histogram of the total number of steps taken each day:

```{r}
total.steps.per.day <- activities.data.table[, sum(steps, na.rm = TRUE), by = date]
with(list(d = total.steps.per.day), {
    d[, total.steps := V1]; d[, V1 := NULL]
    plot(total.steps ~ date, data = d, type = "h",
         main = "The total number of steps taken each day",
         xlab = "Date", ylab = "Total steps", col = "red")
})
```

Mean of the total number of steps taken per day:

```{r}
# Vector of dates without measurements
empty.dates <- activities.data.table[, all(is.na(steps)), by = date][V1 == T][["date"]]
# Excluding empty dates
normal.total.steps.per.day <- total.steps.per.day[!(date %in% empty.dates)]
mean.total.steps.per.day <- mean(normal.total.steps.per.day[["total.steps"]])
print(mean.total.steps.per.day)
```

Median of the total number of steps taken per day:

```{r}
# Excluding empty dates
median.total.steps.per.day <- median(normal.total.steps.per.day[["total.steps"]])
print(median.total.steps.per.day)
```

## What is the average daily activity pattern?

The 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avg.steps.number.per.interval <- activities.data.table[, mean(steps, na.rm = T), by = interval]
with(list(d = avg.steps.number.per.interval), {
    d[, total.steps := V1]
    plot(total.steps ~ interval, data = d, type = "l",
         main = "Average number of steps per interval (avg-ed across all days)",
         xlab = "5-minute interval", ylab = "Avg. steps number")
})
```


Interval, on average across all the days in the dataset, contains the maximum number of steps:

```{r}
with(list(d = avg.steps.number.per.interval), {
    max.interval.row <- d[which.max(d[["total.steps"]])]
    print(max.interval.row)
})

```

## Imputing missing values

Total number of missing values in the dataset:

```{r}
total.na.count <- sum(is.na(activities.data.table[["steps"]]))
sprintf("%.0f%%", total.na.count * 100 / nrow(activities.data.table))
```

Total number of days with some missing values:

```{r}
sum(activities.data.table[, any(is.na(steps)), by = date][["V1"]])
```

Total number of days without measurements at all:

```{r}
sum(activities.data.table[, all(is.na(steps)), by = date][["V1"]])
```

The whole days are missing. That's why the *average number of steps per day* will be appropriate estimation for missing values:

*Before imputing:*

```{r}
summary(activities.data.table)
```

*After imputing:*

```{r}
intervals.per.day <- activities.data.table[, .N, by = date][1][["N"]]
# Estimation
mean.total.steps.per.interval <- mean.total.steps.per.day / intervals.per.day

# Corrected/imputed activities
activities.imputed <- data.table(activities.data.table)
invisible({
    activities.imputed[, steps := as.numeric(steps)]
    activities.imputed[is.na(steps), steps := mean.total.steps.per.interval]
})
summary(activities.imputed)
```

*Imputed data histogram* of the total number of steps taken each day *compared* to actual data:

```{r}
total.steps.per.day.imputed <- activities.imputed[, sum(steps), by = date]
invisible({
    with(list(d = total.steps.per.day.imputed), {
        d[, total.steps := V1]
        d[, V1 := NULL]
        d[, imputed := T]
    })
    total.steps.per.day[, imputed := FALSE]
})
with(list(d = rbind(total.steps.per.day.imputed, total.steps.per.day)), {
    #g <- ggplot(d, aes(date, total.steps, color = imputed)) +
    #    geom_histogram(d[imputed == F], fill = "red", alpha = 0.3) +
    #    geom_histogram(d[imputed == T], fill = "green", alpha = 0.3)
    #print(g)
})
```

*Imputed data mean* of the total number of steps taken per day *compared* to actual value:

```{r}
sprintf("Imputed vs Actual: %.1f vs %.1f", mean(total.steps.per.day.imputed[["total.steps"]]),
        mean.total.steps.per.day)
```

So the mean has not changed ... as expected.

*Imputed data median* of the total number of steps taken per day *compared* to actual value:

```{r}
sprintf("Imputed vs Actual: %.1f vs %.1f", median(total.steps.per.day.imputed[["total.steps"]]),
        median.total.steps.per.day)
```

## Are there differences in activity patterns between weekdays and weekends?
